# Upcoming title

## Synopsis


## Data Processing

### Load libraries 

```{r}

library(stringr)
library(dplyr)

```


### Load data into R

```{r }

storms_df <- read.csv('repdata_data_StormData.csv.bz2', na.strings = c('NA', ''),
                      stringsAsFactors = FALSE)

```

### Examine raw data

```{r}

dim(storms_df)

sum(complete.cases(storms_df))

names(storms_df)

str(storms_df)

```


We take a close look at the `EVTYPE` variable which is extremely important for answering both questions addressed in this report. 

```{r}

unique_raw_events <- length(unique(storms_df$EVTYPE))
unique_raw_events

table(storms_df$EVTYPE, useNA = "ifany")

```


### Pre-process data

First, we begin with a simple pre-processing step by converting all entries in the `EVTYPE` column to lower case and by removing leading and trailing white spaces 

```{r}

storms_df$EVTYPE <- tolower(storms_df$EVTYPE)

length(unique(storms_df$EVTYPE))

storms_df$EVTYPE <- str_trim(storms_df$EVTYPE)

paste0("Total number of unique event types after first two pre-processing steps: " length(unique(storms_df$EVTYPE).

```

The number of unique event types is still too big compared with the 48 values described in the National Weather Service Storm Data Documentation (page 6)


```{r}

```




## Results

Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Across the United States, which types of events have the greatest economic consequences?
